---
title: "hackaton_staa" 
author: "dawnstaa"
date: "4/13/2022"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.asp = 0.618, collapse=TRUE)
```

### Loading needed libraries
```{r message=FALSE}
#Libraries
library(tidyverse)
library(raster)
library(mapdata)  
library(marmap)
library(sf)
```

###  Reading in csv for 2018 and 2016 

```{r}
#using read_csv removes dots from column names
imp_2018 = read_csv("data/2018303d_final.csv") # SC Impaired Waters List 303d for 2018
imp_2016 = read_csv("data/SC_303d_lists_2006to2016/PN_2016303d_final.csv")
#summary(imp_2018) 
#glimpse(imp_2018)
#summary(imp_2016)
#glimpse(imp_2016)
```

### Getting rid of empty rows in 2018 priority rank

```{r}
##2018
dim(imp_2018) #1979 rows
imp_2018 = imp_2018 %>%
  filter(`PRIORITY RANK` != "")#%>%
#filter(`PRIORITY RANK` == "1")#
dim(imp_2018) # 1979-861 = 1118 rows
#glimpse(imp_2018)

##2016
#Getting rid of empty rows in 2016 priority rank
dim(imp_2016) #1503 rows
imp_2016 = imp_2016 %>%
  filter(`PRIORITY RANK` != "") 
dim(imp_2016) # 1979-535 = 1038 rows 
#glimpse(imp_2016)
```

### 2018

```{r}
#Selecting needed columns
imp_2018_use = imp_2018 %>%
  dplyr::select(`PRIORITY RANK`, USE, HUC_12, `CAUSE(S)`)

#glimpse(imp_2018_use)
#summary(imp_2018_use)
```

### Bar plots for priority rank, use and cause - 2018

```{r}
#excluding the 4 rows with multiple ranks
imp_2018_use_no_multipleranks = imp_2018_use %>% 
  slice(-c(128:131))
```

### Bar plot for use and priority rank - 2018

```{r}
bar_2018_rankuse = ggplot(imp_2018_use_no_multipleranks, aes(x=USE, fill= `PRIORITY RANK`)) +
  geom_bar() + #stat = "identity")
  ggtitle("Plot of Priority Rank and Use for 2018")

bar_2018_rankuse
```

### Bar plot for priority rank and cause - 2018

```{r}
bar_2018_rankcause = ggplot(imp_2018_use_no_multipleranks, aes(x=`CAUSE(S)`, fill= `PRIORITY RANK`)) +
  geom_bar() +#stat = "identity") +
  ggtitle("Plot of Priority Rank and Cause for 2018")+
  #facet_wrap(~USE)+
  coord_flip()

bar_2018_rankcause
```

### Bar plot for use and cause - 2018

```{r}
bar_2018_usecause = ggplot(imp_2018_use_no_multipleranks, aes(x=`CAUSE(S)`, fill= USE)) +
  geom_bar() +#stat = "identity") +
  ggtitle("Plot of Use and Cause for 2018")+
  #facet_wrap(~USE)+
  coord_flip()

bar_2018_usecause
```

## 2016 

```{r}
#Selecting needed columns
imp_2016_use = imp_2016 %>%
  dplyr::select(`PRIORITY RANK`, USE, HUC_12, `CAUSE(S)`) 
#glimpse(imp_2018_use)
#summary(imp_2018_use)
```

### Bar plots for priority rank, use and cause - 2016

```{r}
#excluding the 4 rows with multiple ranks
imp_2016_use_no_multipleranks = imp_2016_use %>%
  slice(-c(115,129,137,138,287,288,563,952))
```

### Bar plot for use and priority rank - 2016

```{r}
bar_2016_rankuse = ggplot(imp_2016_use_no_multipleranks, aes(x=USE, fill= `PRIORITY RANK`)) +
  geom_bar() + #stat = "identity")
  ggtitle("Plot of Priority Rank and Use for 2016")

bar_2016_rankuse
```

### Bar plot for priority rank and cause - 2016

```{r}
bar_2016_rankcause = ggplot(imp_2016_use_no_multipleranks, aes(x=`CAUSE(S)`, fill= `PRIORITY RANK`)) +
  geom_bar() +
  ggtitle("Plot of Priority Rank and Cause for 2016")+
  #facet_wrap(~USE)+
  coord_flip()

bar_2016_rankcause
```

### Bar plot for use and cause - 2016

```{r}
bar_2016_usecause = ggplot(imp_2016_use_no_multipleranks, aes(x=`CAUSE(S)`, fill= USE)) +
  geom_bar() +
  ggtitle("Plot of Use and Cause for 2016")+
  #facet_wrap(~USE)+
  coord_flip()

bar_2016_usecause
```










